---
title: "Phylogeny Informed Subsampling Source Analysis"
author: "Emily Dolson"
date: "2023-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

First, we load in the data and create a column containing the number of fitness function evaluations per generation across the different evaluation modes.

```{r setup_data}
library(readr)
library(ggplot2)
library(dplyr)

phylo_informed_all_phylo_data <- read_csv("phylo_informed_all_phylo_data.csv")
phylo_informed_all_phylo_data <- phylo_informed_all_phylo_data %>% mutate(evals_per_gen = case_when(
    EVAL_MODE == "cohort" ~ 1.0/NUM_COHORTS, 
    EVAL_MODE=="down-sample" ~ TEST_DOWNSAMPLE_RATE, 
    EVAL_MODE=="full"~1.0)
  )

```

# Relative-bassed estimation

The first question is where estimated scores are most commonly being pulled from under relative-based estimation. 

Proportion of estimations where the estimate was pulled from the focal taxon itself:

```{r relative_self}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE == "relative-opt", EVAL_MODE!="full")) + geom_boxplot(aes(x=as.factor(evals_per_gen), y=self_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), color=EVAL_MODE, group=as.factor(evals_per_gen))) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(EVAL_MODE)) + theme_bw() + scale_y_continuous("Proportion of self lookups") + scale_x_discrete() + theme(legend.position = "none") + xlab("Evaluations per generation")

```

Proportion of estimations where the estimate was pulled from an ancestor:

```{r relative_ancestor}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE == "relative-opt", EVAL_MODE!="full")) + geom_boxplot(aes(x=as.factor(evals_per_gen), y=ancestor_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), color=EVAL_MODE, group=as.factor(evals_per_gen))) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(EVAL_MODE)) + theme_bw() + scale_y_continuous("Proportion of ancestor lookups") + scale_x_discrete() + theme(legend.position = "none") + xlab("Evaluations per generation")

```

Proportion of estimations where the estimate was pulled from an descendant:

```{r relative_descendant}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE == "relative-opt", EVAL_MODE!="full")) + geom_boxplot(aes(x=as.factor(evals_per_gen), y=descendant_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), color=EVAL_MODE, group=as.factor(evals_per_gen))) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(EVAL_MODE)) + theme_bw() + scale_y_continuous("Proportion of descendant lookups") + scale_x_discrete() + theme(legend.position = "none") + xlab("Evaluations per generation")

```

Proportion of estimations where the estimate was pulled from a non-ancestor, non-descendant relative:

```{r relative_relative}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE == "relative-opt", EVAL_MODE!="full")) + geom_boxplot(aes(x=as.factor(evals_per_gen), y=other_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), color=EVAL_MODE, group=as.factor(evals_per_gen))) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(EVAL_MODE)) + theme_bw() + scale_y_continuous("Proportion of relative lookups") + scale_x_discrete() + theme(legend.position = "none") + xlab("Evaluations per generation")

```

Proportion of estimations where the estimate was pulled from taxon that has since been pruned from the phylogeny (these should all be either relatives or descendants):

```{r relative_outside}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE == "relative-opt", EVAL_MODE!="full")) + geom_boxplot(aes(x=as.factor(evals_per_gen), y=outside_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), color=EVAL_MODE, group=as.factor(evals_per_gen))) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(EVAL_MODE)) + theme_bw() + scale_y_continuous("Proportion of outside lookups") + scale_x_discrete() + theme(legend.position = "none") + xlab("Evaluations per generation")

```

# Comparison between ancesstor-based and relative-based estimation

It appears that down-sampling is only ever drawing estimations from self or ancestors. However, relative-based estimation outperforms ancestor-based for down-sampled lexicase. How?

Let's look at down-sampled lexicase and compare relative-based and ancestor-based estimation. First, the proportion of times an ancestor was used:

```{r relative_vs_ancestor_ancestor}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE != "none", EVAL_MODE=="down-sample")) + geom_boxplot(aes(x=EVAL_FIT_EST_MODE, y=ancestor_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), group=EVAL_FIT_EST_MODE)) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(evals_per_gen)) + theme_bw() + scale_y_continuous("Proportion of ancestor lookups") + scale_x_discrete() + theme(legend.position = "none")
```

Self is just the inverse, but let's double check:

```{r relative_vs_ancestor_slef}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE != "none", EVAL_MODE=="down-sample")) + geom_boxplot(aes(x=EVAL_FIT_EST_MODE, y=self_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), group=EVAL_FIT_EST_MODE)) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(evals_per_gen)) + theme_bw() + scale_y_continuous("Proportion of self lookups") + scale_x_discrete() + theme(legend.position = "none")
```
