---
title: "Phylogeny Informed Subsampling Source Analysis"
author: "Emily Dolson"
date: "2023-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

First, we load in the data and create a column containing the number of fitness function evaluations per generation across the different evaluation modes.

```{r setup_data}
library(readr)
library(ggplot2)
library(dplyr)
```


```{r}
experiment_slug <- "2023-05-07-diag-opt"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

```{r}

phylo_informed_all_phylo_data <- read_csv(
  paste0(working_directory, "data/phylo-est-info.csv")
)

phylo_informed_all_phylo_data <- phylo_informed_all_phylo_data %>%
  mutate(
    evals_per_gen = case_when(
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0
    )
  )

```

# Relative-bassed estimation

The first question is where estimated scores are most commonly being pulled from under relative-based estimation.

Proportion of estimations where the estimate was pulled from the focal taxon itself:

```{r relative_self}
ggplot(
  phylo_informed_all_phylo_data %>%
    filter(EVAL_MODE!="full")
  ) +
  geom_boxplot(
    aes(
      x = as.factor(EVAL_FIT_EST_MODE),
      y = self_count /(other_count+ancestor_count+descendant_count+self_count+outside_count),
      color = EVAL_MODE,
      # group=as.factor(evals_per_gen)
    )
  ) +
  facet_grid(
    cols = vars(DIAGNOSTIC),
    rows = vars(EVAL_MODE)
  ) +
  theme_bw() +
  scale_y_continuous("Proportion of self lookups") +
  scale_x_discrete() +
  theme(legend.position = "none") + xlab("Evaluations per generation")
ggsave(
   filename=paste0(plot_directory, "self-lookups.pdf")
)
```

Proportion of estimations where the estimate was pulled from an ancestor:

```{r relative_ancestor}
ggplot(
  phylo_informed_all_phylo_data %>%
    filter(EVAL_MODE!="full")
  ) +
  geom_boxplot(
    aes(
      x = as.factor(EVAL_FIT_EST_MODE),
      y = ancestor_count /(other_count+ancestor_count+descendant_count+self_count+outside_count),
      color = EVAL_MODE,
      # group=as.factor(evals_per_gen)
    )
  ) +
  facet_grid(
    cols = vars(DIAGNOSTIC),
    rows = vars(EVAL_MODE)
  ) +
  theme_bw() +
  scale_y_continuous("Proportion of ancestor lookups") +
  scale_x_discrete() +
  theme(legend.position = "none") + xlab("Evaluations per generation")
ggsave(
   filename=paste0(plot_directory, "ancestor-lookups.pdf")
)
```

Proportion of estimations where the estimate was pulled from an descendant:

```{r relative_descendant}
ggplot(
  phylo_informed_all_phylo_data %>%
    filter(EVAL_MODE!="full")
  ) +
  geom_boxplot(
    aes(
      x = as.factor(EVAL_FIT_EST_MODE),
      y = descendant_count /(other_count+ancestor_count+descendant_count+self_count+outside_count),
      color = EVAL_MODE,
      # group=as.factor(evals_per_gen)
    )
  ) +
  facet_grid(
    cols = vars(DIAGNOSTIC),
    rows = vars(EVAL_MODE)
  ) +
  theme_bw() +
  scale_y_continuous("Proportion of descendant lookups") +
  scale_x_discrete() +
  theme(legend.position = "none") + xlab("Evaluations per generation")
ggsave(
   filename=paste0(plot_directory, "descendant-lookups.pdf")
)
```

Proportion of estimations where the estimate was pulled from a non-ancestor, non-descendant relative:

```{r relative_relative}
ggplot(
  phylo_informed_all_phylo_data %>%
    filter(EVAL_MODE!="full")
  ) +
  geom_boxplot(
    aes(
      x = as.factor(EVAL_FIT_EST_MODE),
      y = other_count /(other_count+ancestor_count+descendant_count+self_count+outside_count),
      color = EVAL_MODE,
      # group=as.factor(evals_per_gen)
    )
  ) +
  facet_grid(
    cols = vars(DIAGNOSTIC),
    rows = vars(EVAL_MODE)
  ) +
  theme_bw() +
  scale_y_continuous("Proportion of other lookups") +
  scale_x_discrete() +
  theme(legend.position = "none") + xlab("Evaluations per generation")
ggsave(
   filename=paste0(plot_directory, "other-lookups.pdf")
)
```

Proportion of estimations where the estimate was pulled from taxon that has since been pruned from the phylogeny (these should all be either relatives or descendants):

```{r relative_outside}
ggplot(
  phylo_informed_all_phylo_data %>%
    filter(EVAL_MODE!="full")
  ) +
  geom_boxplot(
    aes(
      x = as.factor(EVAL_FIT_EST_MODE),
      y = outside_count /(other_count+ancestor_count+descendant_count+self_count+outside_count),
      color = EVAL_MODE,
      # group=as.factor(evals_per_gen)
    )
  ) +
  facet_grid(
    cols = vars(DIAGNOSTIC),
    rows = vars(EVAL_MODE)
  ) +
  theme_bw() +
  scale_y_continuous("Proportion of outside lookups") +
  scale_x_discrete() +
  theme(legend.position = "none") + xlab("Evaluations per generation")
ggsave(
   filename=paste0(plot_directory, "outside-lookups.pdf")
)
```

# Comparison between ancesstor-based and relative-based estimation

It appears that down-sampling is only ever drawing estimations from self or ancestors. However, relative-based estimation outperforms ancestor-based for down-sampled lexicase. How?

Let's look at down-sampled lexicase and compare relative-based and ancestor-based estimation. First, the proportion of times an ancestor was used:

```{r relative_vs_ancestor_ancestor}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE != "none", EVAL_MODE=="down-sample")) + geom_boxplot(aes(x=EVAL_FIT_EST_MODE, y=ancestor_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), group=EVAL_FIT_EST_MODE)) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(evals_per_gen)) + theme_bw() + scale_y_continuous("Proportion of ancestor lookups") + scale_x_discrete() + theme(legend.position = "none")
```

Self is just the inverse, but let's double check:

```{r relative_vs_ancestor_slef}
ggplot(phylo_informed_all_phylo_data %>% filter(EVAL_FIT_EST_MODE != "none", EVAL_MODE=="down-sample")) + geom_boxplot(aes(x=EVAL_FIT_EST_MODE, y=self_count/(other_count+ancestor_count+descendant_count+self_count+outside_count), group=EVAL_FIT_EST_MODE)) + facet_grid(cols=vars(DIAGNOSTIC), rows=vars(evals_per_gen)) + theme_bw() + scale_y_continuous("Proportion of self lookups") + scale_x_discrete() + theme(legend.position = "none")
```
