# 2023-12-28 Analysis

```{r}
experiment_slug <- "2023-12-28-phylo-sampling-diag"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)

if (exists("bookdown_wd_prefix")) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}
```

## Dependencies

```{r}
library(tidyverse)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
library()
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)

full_median_size = 1.5
```

### Load experiment summary data

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

summary_data <- summary_data %>%
  mutate(
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0,
      EVAL_MODE == "indiv-rand-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "phylo-informed-sample" ~ TEST_DOWNSAMPLE_RATE
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    eval_label = case_when(
      EVAL_MODE == "down-sample" & EVAL_FIT_EST_MODE != "none" ~ paste("down-sample", EVAL_FIT_EST_MODE, sep="-"),
      .default = EVAL_MODE
    )
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

# Split summary data on diagnostic
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)

explore_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "multipath-exploration"
)

exploit_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "exploitation-rate"
)
```

### Load experiment time series data

```{r}
ts_data_loc <- paste0(working_directory, "data/time_series.csv")
ts_data <- read_csv(ts_data_loc)

ts_data <- ts_data %>%
  mutate(
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0,
      EVAL_MODE == "indiv-rand-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "phylo-informed-sample" ~ TEST_DOWNSAMPLE_RATE
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    eval_label = case_when(
      EVAL_MODE == "down-sample" & EVAL_FIT_EST_MODE != "none" ~ paste("down-sample", EVAL_FIT_EST_MODE, sep="-"),
      .default = EVAL_MODE
    )
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

con_obj_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives")

explore_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "multipath-exploration")

exploit_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "exploitation-rate")
```

Summarize time series data

```{r}
ts_summary_data <- ts_data %>%
  group_by(SEED, DIAGNOSTIC, SELECTION, evals_per_gen, eval_label) %>%
  summarize(
    n = n(),
    avg_num_unique_selected = mean(num_unique_selected),
    total_optimal_trait_coverage_loss = sum(optimal_trait_coverage_loss)
  )
```

## Contradictory objectives diagnostic

### Population-wide satisfactory trait coverage (final)

Satisfactory trait coverage after 50,000 generations

```{r}

build_con_obj_cov_f_plot <- function(selection) {

  full_median <- median(
    filter(
      con_obj_summary_data,
      eval_label == "full" & SELECTION == selection
    )$pop_optimal_trait_coverage
  )

  plot <- con_obj_summary_data %>%
    filter(
      eval_label != "full" & SELECTION == selection
    ) %>%
    ggplot(
      aes(
        x = eval_label,
        y = pop_optimal_trait_coverage,
        fill = eval_label
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust = 1.5
    ) +
    geom_point(
      mapping = aes(color = eval_label),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      # limits = c(-0.5, 100)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_grid(
      SELECTION ~ evals_per_gen,
      # nrow=2,
      labeller = label_both
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )

  return(plot)

}

# tourn_co_plt <- build_con_obj_cov_f_plot("tournament")
lex_co_plt <- build_con_obj_cov_f_plot("lexicase")

ggsave(
  filename = paste0(plot_directory, "contra-obj-cov-final-lex.pdf"),
  plot = lex_co_plt + labs(title = "Contradictory objectives"),
  width = 15,
  height = 10
)
```

### Population-wide satisfactory trait coverage (over time)

```{r}
con_obj_pop_cov_ts <- con_obj_ts_data %>%
  filter(SELECTION == "lexicase") %>%
  ggplot(
    aes(
      x = ts_step,
      y = pop_optimal_trait_coverage,
      fill = eval_label,
      color = eval_label
    )
  ) +
  stat_summary(
    geom = "line",
    fun = mean
  ) +
  stat_summary(
    geom = "ribbon",
    fun.data = "mean_cl_boot",
    fun.args = list(conf.int = 0.95),
    alpha = 0.2,
    linetype = 0
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    ~ evals_per_gen,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-ts.pdf"),
  plot = con_obj_pop_cov_ts + labs(title="Contradictory objectives"),
  width = 10,
  height = 15
)
```

### MRCA changes (successful selective sweeps)

```{r}

full_median <- median(
  filter(
    con_obj_summary_data,
    eval_label == "full" & SELECTION == "lexicase"
  )$phylo_mrca_changes
)

plot <- con_obj_summary_data %>%
  filter(
    eval_label != "full" & SELECTION == "lexicase"
  ) %>%
  ggplot(
    aes(
      x = eval_label,
      y = phylo_mrca_changes,
      fill = eval_label
    )
  ) +
  geom_hline(
    yintercept = full_median,
    size = full_median_size,
    alpha = 0.7,
    color = "black"
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust = 1.5
  ) +
  geom_point(
    mapping = aes(color = eval_label),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    SELECTION ~ evals_per_gen,
    # nrow=2,
    labeller = label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-mrca-chgs-final-lex.pdf"),
  plot = plot + labs(title = "Contradictory objectives"),
  width = 15,
  height = 10
)
```

### Mean genotype deleterious steps (over time)
<!-- mean_genoetype_deleterious_steps -->

```{r}
plot <- con_obj_ts_data %>%
  filter(SELECTION == "lexicase") %>%
  ggplot(
    aes(
      x = ts_step,
      y = mean_genoetype_deleterious_steps,
      fill = eval_label,
      color = eval_label
    )
  ) +
  stat_summary(
    geom = "line",
    fun = mean
  ) +
  stat_summary(
    geom = "ribbon",
    fun.data = "mean_cl_boot",
    fun.args = list(conf.int = 0.95),
    alpha = 0.2,
    linetype = 0
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    ~ evals_per_gen,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-ts-del-steps.pdf"),
  plot = plot + labs(title="Contradictory objectives"),
  width = 10,
  height = 15
)
```

### Mean genotype deleterious steps (final)

```{r}
full_median <- median(
  filter(
    con_obj_summary_data,
    eval_label == "full" & SELECTION == "lexicase"
  )$phylo_mean_genoetype_deleterious_steps
)

plot <- con_obj_summary_data %>%
  filter(
    eval_label != "full" & SELECTION == "lexicase"
  ) %>%
  ggplot(
    aes(
      x = eval_label,
      y = phylo_mean_genoetype_deleterious_steps,
      fill = eval_label
    )
  ) +
  geom_hline(
    yintercept = full_median,
    size = full_median_size,
    alpha = 0.7,
    color = "black"
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust = 1.5
  ) +
  geom_point(
    mapping = aes(color = eval_label),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    SELECTION ~ evals_per_gen,
    # nrow=2,
    labeller = label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-del-steps-final-lex.pdf"),
  plot = plot + labs(title = "Contradictory objectives"),
  width = 15,
  height = 10
)
```

### Mean genotype pairwise distance (over time)
<!-- mean_genotype_pairwise_distance -->
```{r}
plot <- con_obj_ts_data %>%
  filter(SELECTION == "lexicase") %>%
  ggplot(
    aes(
      x = ts_step,
      y = mean_genotype_pairwise_distance,
      fill = eval_label,
      color = eval_label
    )
  ) +
  stat_summary(
    geom = "line",
    fun = mean
  ) +
  stat_summary(
    geom = "ribbon",
    fun.data = "mean_cl_boot",
    fun.args = list(conf.int = 0.95),
    alpha = 0.2,
    linetype = 0
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    ~ evals_per_gen,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-ts-pw-dist.pdf"),
  plot = plot + labs(title="Contradictory objectives"),
  width = 10,
  height = 15
)
```

### Mean genotype pairwise distance (final)
```{r}
full_median <- median(
  filter(
    con_obj_summary_data,
    eval_label == "full" & SELECTION == "lexicase"
  )$phylo_mean_genotype_pairwise_distance
)

plot <- con_obj_summary_data %>%
  filter(
    eval_label != "full" & SELECTION == "lexicase"
  ) %>%
  ggplot(
    aes(
      x = eval_label,
      y = phylo_mean_genotype_pairwise_distance,
      fill = eval_label
    )
  ) +
  geom_hline(
    yintercept = full_median,
    size = full_median_size,
    alpha = 0.7,
    color = "black"
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust = 1.5
  ) +
  geom_point(
    mapping = aes(color = eval_label),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    SELECTION ~ evals_per_gen,
    # nrow=2,
    labeller = label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-pw-dist-final-lex.pdf"),
  plot = plot + labs(title = "Contradictory objectives"),
  width = 15,
  height = 10
)
```

### Number unique individual selected (averaged over all generations)
<!-- num_unique_selected -->

```{r}
full_median <- median(
  filter(
    ts_summary_data,
    DIAGNOSTIC == "contradictory-objectives" && eval_label == "full" & SELECTION == "lexicase"
  )$avg_num_unique_selected
)

plot <- ts_summary_data %>%
  filter(
    DIAGNOSTIC == "contradictory-objectives" & eval_label != "full" & SELECTION == "lexicase"
  ) %>%
  ggplot(
    aes(
      x = eval_label,
      y = avg_num_unique_selected,
      fill = eval_label
    )
  ) +
  geom_hline(
    yintercept = full_median,
    size = full_median_size,
    alpha = 0.7,
    color = "black"
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust = 1.5
  ) +
  geom_point(
    mapping = aes(color = eval_label),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    SELECTION ~ evals_per_gen,
    # nrow=2,
    labeller = label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-avg-unique-sel-lex.pdf"),
  plot = plot + labs(title = "Contradictory objectives"),
  width = 15,
  height = 10
)
```

## Exploration

### Max aggregate score (final)

```{r}
build_plot_summary_data <- function(diagnostic, selection, response) {
  diag_data <- summary_data %>% filter(DIAGNOSTIC == diagnostic)

  full_median <- median(
    filter(
      diag_data,
      eval_label == "full" & SELECTION == selection
    )[[response]]
  )

  plot <- diag_data %>%
    filter(
      eval_label != "full" & SELECTION == selection
    ) %>%
    ggplot(
      aes_string(
        x = "eval_label",
        y = response,
        fill = "eval_label"
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust = 1.5
    ) +
    geom_point(
      mapping = aes(color = eval_label),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      # limits = c(-0.5, 100)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_grid(
      SELECTION ~ evals_per_gen,
      # nrow=2,
      labeller = label_both
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )

  return(plot)
}

# tourn_co_plt <- build_con_obj_cov_f_plot("tournament")
p <- build_plot_summary_data("multipath-exploration", "lexicase", "elite_true_agg_score")

ggsave(
  filename = paste0(plot_directory, "explore-score-final-lex.pdf"),
  plot = p + labs(title = "Exploration"),
  width = 15,
  height = 10
)
```

### Max aggregate score (over time)

```{r}
build_plot_time_series <- function(diagnostic, selection, response) {
  diag_data <- ts_data %>% filter(DIAGNOSTIC == diagnostic)

  plot <- diag_data %>%
    filter(
      SELECTION == selection
    ) %>%
    ggplot(
      aes_string(
        x = "ts_step",
        y = {{ response }},
        fill = "eval_label",
        color = "eval_label"
      )
    ) +
    stat_summary(
      geom = "line",
      fun = mean
    ) +
    stat_summary(
      geom = "ribbon",
      fun.data = "mean_cl_boot",
      fun.args = list(conf.int = 0.95),
      alpha = 0.2,
      linetype = 0
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
      ~ evals_per_gen,
      ncol = 1,
      labeller = label_both
    ) +
    theme(
      legend.position = "bottom"
    )

  return(plot)
}

p <- build_plot_time_series("multipath-exploration", "lexicase", "max_agg_score")
ggsave(
  filename = paste0(plot_directory, "explore-score-ts-lex.pdf"),
  plot = p + labs(title = "Exploration"),
  width = 15,
  height = 10
)
```

## Exploitation

### Aggregate score (final)

```{r}

p <- build_plot_summary_data("exploitation-rate", "lexicase", "elite_true_agg_score")
ggsave(
  filename = paste0(plot_directory, "exploit-score-final-lex.pdf"),
  plot = p + labs(title = "Exploitation rate"),
  width = 15,
  height = 10
)
```

```{r}

p <- build_plot_summary_data("exploitation-rate", "tournament", "elite_true_agg_score")
ggsave(
  filename = paste0(plot_directory, "exploit-score-final-tourn.pdf"),
  plot = p + labs(title = "Exploitation rate"),
  width = 15,
  height = 10
)
```

### Aggregate score (over time)

```{r}

p <- build_plot_time_series("exploitation-rate", "lexicase", "max_agg_score")
ggsave(
  filename = paste0(plot_directory, "exploit-score-ts-lex.pdf"),
  plot = p + labs(title = "Exploitation rate"),
  width = 15,
  height = 10
)
```

```{r}

p <- build_plot_time_series("exploitation-rate", "tournament", "max_agg_score")
ggsave(
  filename = paste0(plot_directory, "exploit-score-ts-tourn.pdf"),
  plot = p + labs(title = "Exploitation rate"),
  width = 15,
  height = 10
)
```



