# 2023-12-28 - Contradictory objectives diagnostics

```{r}
experiment_slug <- "2023-12-28-phylo-sampling-diag"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)

if (exists("bookdown_wd_prefix")) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}
```

## Dependencies

```{r}
library(tidyverse)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
# Constants
focal_diagnostic <- "contradictory-objectives"
```

### Load experiment summary data

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

summary_data <- summary_data %>%
  mutate(
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0,
      EVAL_MODE == "indiv-rand-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "phylo-informed-sample" ~ TEST_DOWNSAMPLE_RATE
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    eval_label = case_when(
      # Clean up down-sample label
      EVAL_MODE == "down-sample" & EVAL_FIT_EST_MODE != "none" ~ paste("down-sample", EVAL_FIT_EST_MODE, sep="-"),
      .default = EVAL_MODE
    ),
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

# Grab just the contradictory objectives data
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)
```

### Load experiment time series data

```{r}
ts_data_loc <- paste0(working_directory, "data/time_series.csv")
ts_data <- read_csv(ts_data_loc)

ts_data <- ts_data %>%
  mutate(
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0,
      EVAL_MODE == "indiv-rand-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "phylo-informed-sample" ~ TEST_DOWNSAMPLE_RATE
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    eval_label = case_when(
      EVAL_MODE == "down-sample" & EVAL_FIT_EST_MODE != "none" ~ paste("down-sample", EVAL_FIT_EST_MODE, sep="-"),
      .default = EVAL_MODE
    )
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

# Grab just the contradictory objectives data
con_obj_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives")

```

Summarize time series data

```{r}
ts_summary_data <- ts_data %>%
  group_by(SEED, DIAGNOSTIC, SELECTION, evals_per_gen, eval_label) %>%
  summarize(
    n = n(),
    avg_num_unique_selected = mean(num_unique_selected),
    total_optimal_trait_coverage_loss = sum(optimal_trait_coverage_loss)
  )
```

### Plotting helper functions

Functions to assist with quick plotting of different measurements from summary and timeseries data.

```{r}
build_plot_summary_data <- function(data, diagnostic, selection, response) {
  diag_data <- data %>% filter(DIAGNOSTIC == diagnostic)

  full_median <- median(
    filter(
      diag_data,
      eval_label == "full" & SELECTION == selection
    )[[response]]
  )

  plot <- diag_data %>%
    filter(
      eval_label != "full" & SELECTION == selection
    ) %>%
    ggplot(
      aes_string(
        x = "eval_label",
        y = response,
        fill = "eval_label"
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = 1.0,
      alpha = 0.7,
      color = "black",
      linetype="dashed"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust = 1.5
    ) +
    geom_point(
      mapping = aes(color = eval_label),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      # limits = c(-0.5, 100)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_grid(
      SELECTION ~ evals_per_gen,
      # nrow=2,
      labeller = label_both
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color = "gray", size = 2)
    )

  return(plot)
}

build_plot_time_series_single_sampling <- function(
  data,
  diagnostic,
  selection,
  sampling_level,
  response
) {

  diag_data <- data %>% filter(
    DIAGNOSTIC == diagnostic &
    SELECTION == selection &
    evals_per_gen == sampling_level
  ) %>%
  mutate(
    sampling_level_label = sampling_level
  )

  full_diag_data <- data %>% filter(
    DIAGNOSTIC == diagnostic & SELECTION == selection & eval_label == "full"
  ) %>%
  mutate(
    # Ensure that median line will sit in same facet
    sampling_level_label = sampling_level
  )

  plot <- diag_data %>%
    filter(
      eval_label != "full"
    ) %>%
    ggplot(
      aes_string(
        x = "ts_step",
        # x = "evaluations",
        y = {{ response }}
      )
    ) +
    stat_summary(
      geom = "line",
      fun = mean,
      aes(
        color = eval_label
      )
    ) +
    stat_summary(
      geom = "ribbon",
      fun.data = "mean_cl_boot",
      fun.args = list(conf.int = 0.95),
      alpha = 0.2,
      linetype = 0,
      aes(
        color = eval_label,
        fill = eval_label
      )
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    # facet_wrap(
    #   ~ sampling_level_label,
    #   ncol = 1,
    #   labeller = label_both
    # ) +
    theme(
      legend.position = "right"
    ) +
    stat_summary(
      data = full_diag_data,
      geom = "line",
      fun = median,
      linetype = "dashed",
      color = "black"
    )

  return(plot)
}

build_plot_time_series <- function(
  data,
  diagnostic,
  selection,
  response
) {
  # Build 1% sampling plot and 10% sampling plot
  p_01 <- data  %>% build_plot_time_series_single_sampling(
    diagnostic,
    selection,
    "0.01",
    response
  )
  p_10 <- data %>% build_plot_time_series_single_sampling(
    diagnostic,
    selection,
    "0.1",
    response
  )

  title <- ggdraw() +
    draw_label(
      paste0(diagnostic, " - ", selection),
      fontface = 'bold',
      x = 0,
      hjust = 0
    ) +
    theme(
      # add margin on the left of the drawing canvas,
      # so title is aligned with left edge of first plot
      plot.margin = margin(0, 0, 0, 7)
    )

  plot <- plot_grid(
    title,
    p_01 + labs(title = "1% subsampling") + theme(legend.position = "none"),
    p_10 + labs(title = "10% subsampling") + theme(legend.position = "bottom"),
    nrow = 3,
    ncol = 1,
    rel_heights = c(0.075, 1, 1)
  )

  return(plot)
}
```

## Population-wide satisfactory trait coverage

### Final - Lexicase selection

```{r}
p <- summary_data %>% build_plot_summary_data(
  focal_diagnostic,
  "lexicase",
  "pop_optimal_trait_coverage"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-sat-cov-final-lex.pdf"),
  plot = p + labs(title = "Contradictory objectives - Lexicase selection"),
  width = 15,
  height = 10
)
```

### Over time - lexicase selection

```{r}
p <- ts_data  %>% build_plot_time_series(
  focal_diagnostic,
  "lexicase",
  "pop_optimal_trait_coverage"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-sat-cov-ts-lex.pdf"),
  plot = p,
  width = 15,
  height = 10
)
```

### Final - Tournament selection

```{r}
p <- summary_data %>% build_plot_summary_data(
  focal_diagnostic,
  "tournament",
  "pop_optimal_trait_coverage"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-sat-cov-final-tourn.pdf"),
  plot = p + labs(title = "Contradictory objectives - Tournament selection"),
  width = 15,
  height = 10
)
```

### Over time  - tournament selection

```{r}
p <- ts_data  %>% build_plot_time_series(
  focal_diagnostic,
  "tournament",
  "pop_optimal_trait_coverage"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-sat-cov-ts-tourn.pdf"),
  plot = p,
  width = 15,
  height = 10
)
```

## MRCA changes

```{r}
p <- summary_data %>% build_plot_summary_data(
  focal_diagnostic,
  "lexicase",
  "phylo_mrca_changes"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-mrca-chgs-final-lex.pdf"),
  plot = p + labs(title = "Contradictory objectives - Lexicase selection"),
  width = 15,
  height = 10
)
```

## Mean genotype deleterious steps

```{r}
p <- summary_data %>% build_plot_summary_data(
  focal_diagnostic,
  "lexicase",
  "phylo_mean_genoetype_deleterious_steps"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-del-steps-final-lex.pdf"),
  plot = p + labs(title = "Contradictory objectives - Lexicase selection"),
  width = 15,
  height = 10
)
```

## Mean genotype pairwise distance

```{r}
p <- summary_data %>% build_plot_summary_data(
  focal_diagnostic,
  "lexicase",
  "phylo_mean_genotype_pairwise_distance"
)
ggsave(
  filename = paste0(plot_directory, "con-obj-pw-dist-final-lex.pdf"),
  plot = p + labs(title = "Contradictory objectives - Lexicase selection"),
  width = 15,
  height = 10
)
```

## Number unique individual selected

```{r}
build_plot_summary_data(
  ts_summary_data,
  focal_diagnostic,
  "lexicase",
  "avg_num_unique_selected"
)
```

## Manuscript figures

Time series graphs don't add a ton here, so just final graphs.

```{r}
build_final_score_manuscript_plot <- function(
  selection,
  subsample_rate
) {

  # Extract median values for max aggregate score at same evaluation level as sampling regimes
  max_eval <- max(
    filter(con_obj_summary_data, evals_per_gen == subsample_rate)$evaluations
  )
  full_eval_steps <- as.numeric(
    levels(
      as.factor(
        filter(con_obj_summary_data, eval_label == "full" & evaluations >= max_eval)$evaluations # nolint: line_length_linter.
      )
    )
  )
  full_eval <- full_eval_steps[which.min( full_eval_steps  - max_eval )]
  full_median_score_evals <- median(
    filter(
      con_obj_summary_data,
      SELECTION == selection & eval_label == "full" & evaluations == full_eval
    )$pop_optimal_trait_coverage
  )

  plot <- con_obj_summary_data %>%
    filter(
      eval_label != "full" &
      SELECTION == selection &
      evals_per_gen == subsample_rate
    ) %>%
    ggplot(
      aes(
        x = eval_label,
        y = pop_optimal_trait_coverage,
        fill = eval_label
      )
    ) +
    geom_hline(
      yintercept = full_median_score_evals,
      size = 1.0,
      alpha = 0.7,
      color = "black",
      linetype="dashed"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust = 1.5
    ) +
    geom_point(
      mapping = aes(color = eval_label),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      name = "Satisfactory trait coverage",
      limits = c(0, 100)
    ) +
    scale_x_discrete(
      name = "Subsampling regime",
      breaks = c("down-sample", "down-sample-ancestor", "indiv-rand-sample", "phylo-informed-sample"),
      labels = c("DS", "DS+EST", "IRS", "ABS")
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    theme(
      legend.position = "none",
      # axis.text.x = element_text(
      #   angle = 30,
      #   hjust = 1
      # ),
    )
  return(plot)
}
```

Build end-of-run plots (fixed number of evaluations)

```{r}
plot_final_lex_01 <- build_final_score_manuscript_plot(
  "lexicase",
  "0.01"
)
plot_final_lex_10 <- build_final_score_manuscript_plot(
  "lexicase",
  "0.1"
)
```

Combine into single figure

```{r}
lex_fig <- plot_grid(
  plot_final_lex_01 +
    # labs(
    #   title = "1% subsampling"
    # ) +
    theme(
      plot.margin = margin(1, 0, 0, 0, "cm")
    ),
  plot_final_lex_10 +
    # labs(
    #   title = "10% subsampling"
    # ) +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = margin(1, 0, 0, 1, "cm")
    ),
  nrow = 1,
  ncol = 2,
  align = "h",
  labels = c("a) 1% subsampling", "b) 10% subsampling"),
  rel_widths = c(1, 1)
)
lex_fig
save_plot(
  filename = paste0(plot_directory, "2023-12-28-con-obj-lex-fig.pdf"),
  plot = lex_fig,
  base_width = 7,
  base_height = 4,
  dpi = 600
)
```