# 2023-05-10 - Diagnostics experiments

```{r}
experiment_slug <- "2023-05-10-diagnostics"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)
```

## Dependencies

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

### Load experiment summary data

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

summary_data <- summary_data %>%
  mutate(
    eval_mode_row = case_when(
      EVAL_MODE == "full" & TEST_DOWNSAMPLE_RATE == "1" ~ "down-sample",
      EVAL_MODE == "full" & NUM_COHORTS == "1" ~ "cohort",
      .default = EVAL_MODE
    ),
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    eval_mode_row = as.factor(eval_mode_row),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

# Split summary data on diagnostic
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)

explore_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "multipath-exploration"
)
```

### Load experiment time series data

```{r}
ts_data_loc <- paste0(working_directory, "data/time_series.csv")
ts_data <- read_csv(ts_data_loc)

ts_data <- ts_data %>%
  mutate(
    eval_mode_row = case_when(
      EVAL_MODE == "full" & TEST_DOWNSAMPLE_RATE == "1" ~ "down-sample",
      EVAL_MODE == "full" & NUM_COHORTS == "1" ~ "cohort",
      .default = EVAL_MODE
    ),
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  )

con_obj_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives")

explore_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "multipath-exploration")
```

### Load estimate source distributions

```{r}
est_source_data <- read_csv(
  paste0(working_directory, "data/phylo-est-info.csv")
)

est_source_data <- est_source_data %>%
  mutate(
    eval_mode_row = case_when(
      EVAL_MODE == "full" & TEST_DOWNSAMPLE_RATE == "1" ~ "down-sample",
      EVAL_MODE == "full" & NUM_COHORTS == "1" ~ "cohort",
      .default = EVAL_MODE
    ),
    evals_per_gen = case_when(
      EVAL_MODE == "cohort-full-compete" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "cohort" ~ 1.0 / NUM_COHORTS,
      EVAL_MODE == "down-sample" ~ TEST_DOWNSAMPLE_RATE,
      EVAL_MODE == "full" ~ 1.0
    ),
    EVAL_FIT_EST_MODE = case_when(
      EVAL_FIT_EST_MODE == "ancestor-opt" ~ "ancestor",
      EVAL_FIT_EST_MODE == "relative-opt" ~ "relative",
      .default = EVAL_FIT_EST_MODE
    ),
    .keep = "all"
  ) %>%
  mutate(
    evals_per_gen = as.factor(evals_per_gen),
    eval_mode_row = as.factor(eval_mode_row),
    DIAGNOSTIC = as.factor(DIAGNOSTIC),
    SELECTION = as.factor(SELECTION),
    EVAL_MODE = as.factor(EVAL_MODE),
    NUM_COHORTS = as.factor(NUM_COHORTS),
    TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
    EVAL_FIT_EST_MODE = factor(
      EVAL_FIT_EST_MODE,
      levels = c(
        "none",
        "ancestor",
        "relative"
      ),
      labels = c(
        "None",
        "Ancestor",
        "Relative"
      )
    )
  ) %>%
  mutate(
    prop_self_lookups = case_when(
      (EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") ~
        self_count / (other_count + ancestor_count + descendant_count + self_count + outside_count),
      .default = 0
    ),
    prop_other_lookups =  case_when(
      (EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") ~
        other_count / (other_count + ancestor_count + descendant_count + self_count + outside_count),
      .default = 0
    ),
    prop_ancestor_lookups =  case_when(
      (EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") ~
        ancestor_count / (other_count + ancestor_count + descendant_count + self_count + outside_count),
      .default = 0
    ),
    prop_descendant_lookups = case_when(
      (EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") ~
        descendant_count / (other_count + ancestor_count + descendant_count + self_count + outside_count),
      .default = 0
    ),
    prop_outside_lookups =  case_when(
      (EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") ~
        outside_count / (other_count + ancestor_count + descendant_count + self_count + outside_count),
      .default = 0
    )
  )
```

## Contradictory objectives diagnostic

### Population-wide satisfactory trait coverage (final)

Satifactory trait coverage after 50,000 generations:

```{r}
contradictory_obj_final_plt <- ggplot(
    con_obj_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = pop_optimal_trait_coverage,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    eval_mode_row~evals_per_gen,
    # nrow=2,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color="gray", size=2)
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-final.pdf"),
  plot = contradictory_obj_final_plt + labs(title="Contradictory objectives"),
  width = 15,
  height = 10
)
```

```{r}
contradictory_obj_final_plt
```

#### Statistical analysis

First, we create a table of summary values for satisfactory trait coverage in the final generation.

```{r}
con_obj_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen, EVAL_FIT_EST_MODE) %>%
  summarize(
    cov_median = median(pop_optimal_trait_coverage),
    cov_mean = mean(pop_optimal_trait_coverage),
    n = n()
  ) %>%
  kable()
```

Next, we perform a Kruskal-Wallis test to determine which comparisons contain statistically significant differences among treatments.

```{r}
con_obj_kw_test <- con_obj_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  kruskal_test(pop_optimal_trait_coverage ~ EVAL_FIT_EST_MODE) %>%
  unite(
    "comparison_group",
    EVAL_MODE,
    evals_per_gen,
    sep = "_",
    remove = FALSE
  )

kable(con_obj_kw_test)
```

Finally, we perform a pairwise Wilcoxon rank-sum test (using a Holm-Bonferroni correction for multiple comparisons).
Note that only results from signific

```{r}
sig_kw_groups <- filter(con_obj_kw_test, p < 0.05)$comparison_group

con_obj_stats <- con_obj_summary_data %>%
  unite(
    "comparison_group",
    EVAL_MODE,
    evals_per_gen,
    sep = "_",
    remove = FALSE
  ) %>%
  filter(EVAL_MODE != "full" & comparison_group %in% sig_kw_groups) %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  pairwise_wilcox_test(pop_optimal_trait_coverage ~ EVAL_FIT_EST_MODE) %>%
  adjust_pvalue(method = "holm") %>%
  add_significance("p.adj")

kable(con_obj_stats)

# con_obj_stats %>%
#   filter(p.adj <= 0.05) %>%
#   arrange(
#     desc(p.adj)
#   ) %>%
#   kable()
```

### Population-wide satisfactory trait coverage (over time)

```{r}
contradictory_obj_pop_cov_ts <- ggplot(
    con_obj_ts_data,
    aes(
      x = ts_step,
      y = pop_optimal_trait_coverage,
      fill = EVAL_FIT_EST_MODE,
      color = EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(
    geom = "line",
    fun = mean
  ) +
  stat_summary(
    geom = "ribbon",
    fun.data = "mean_cl_boot",
    fun.args = list(conf.int = 0.95),
    alpha = 0.2,
    linetype = 0
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    EVAL_MODE ~ evals_per_gen,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave(
  filename = paste0(plot_directory, "contra-obj-ts.pdf"),
  plot = contradictory_obj_pop_cov_ts + labs(title="Contradictory objectives"),
  width = 10,
  height = 15
)
```

```{r}
contradictory_obj_pop_cov_ts
```

### Phylogeny estimate source distributions

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_self_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of self lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "contra-obj-self-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_ancestor_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of ancestor lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "contra-obj-ancestor-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_descendant_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of descendant lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "contra-obj-descendant-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_other_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of other lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "contra-obj-other-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_outside_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of outside lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "contra-obj-outside-lookups.pdf")
)
```

## Multi-path exploration diagnostic

### Maximum aggregate score (final)

```{r}
explore_final_score_plt <- ggplot(
    explore_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    eval_mode_row~evals_per_gen,
    # nrow=2,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color="gray", size=2)
  )
ggsave(
  filename = paste0(plot_directory, "explore-final.pdf"),
  plot = explore_final_score_plt + labs(title="Multi-path exploration"),
  width = 15,
  height = 10
)
```

#### Statistical analysis

```{r}
explore_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen, EVAL_FIT_EST_MODE) %>%
  summarize(
    score_median = median(max_agg_score),
    score_mean = mean(max_agg_score),
    n = n()
  ) %>%
  kable()
```

```{r}
explore_kw_test <- explore_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  kruskal_test(max_agg_score ~ EVAL_FIT_EST_MODE) %>%
  mutate(
    sig = (p <= 0.05)
  ) %>%
  unite(
    "comparison_group",
    EVAL_MODE,
    evals_per_gen,
    sep = "_",
    remove = FALSE
  )

kable(explore_kw_test)
```

```{r}
expl_sig_kw_groups <- filter(explore_kw_test, p < 0.05)$comparison_group

explore_stats <- explore_summary_data %>%
  unite(
    "comparison_group",
    EVAL_MODE,
    evals_per_gen,
    sep = "_",
    remove = FALSE
  ) %>%
  filter(EVAL_MODE != "full" & comparison_group %in% expl_sig_kw_groups) %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  pairwise_wilcox_test(max_agg_score ~ EVAL_FIT_EST_MODE) %>%
  adjust_pvalue(method = "holm") %>%
  add_significance("p.adj")

kable(explore_stats)

# explore_stats %>%
#   filter(p.adj <= 0.05) %>%
#   arrange(
#     desc(p.adj)
#   ) %>%
#   kable()
```

### Maximum aggregate score (over time)

```{r}
explore_score_ts <- ggplot(
    explore_ts_data,
    aes(
      x = ts_step,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE,
      color = EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(
    geom = "line",
    fun = mean
  ) +
  stat_summary(
    geom = "ribbon",
    fun.data = "mean_cl_boot",
    fun.args = list(conf.int = 0.95),
    alpha = 0.2,
    linetype = 0
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    EVAL_MODE ~ evals_per_gen,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom"
  )

ggsave(
  filename = paste0(plot_directory, "explore-ts.pdf"),
  plot = explore_score_ts + labs(title="Multi-path exploration"),
  width = 10,
  height = 15
)
```

```{r}
explore_score_ts
```

### Phylogeny estimate source distributions

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "multipath-exploration") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_self_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of self lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "explore-self-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "multipath-exploration") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_ancestor_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of ancestor lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "explore-ancestor-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "multipath-exploration") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_descendant_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of descendant lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "explore-descendant-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "multipath-exploration") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_other_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of other lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "explore-other-lookups.pdf")
)
```

```{r}
est_source_data %>%
  filter(DIAGNOSTIC == "multipath-exploration") %>%
  filter(EVAL_MODE != "full" & EVAL_FIT_EST_MODE != "None") %>%
  ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = prop_outside_lookups
      )
    ) +
    geom_boxplot() +
    geom_point() +
    facet_grid(
      cols = vars(evals_per_gen),
      rows = vars(EVAL_MODE),
      labeller = label_both
    ) +
    scale_y_continuous("Proportion of outside lookups") +
    scale_x_discrete("Evaluations per generation") +
    theme_bw() +
    theme(legend.position = "none")

ggsave(
   filename=paste0(plot_directory, "explore-outside-lookups.pdf")
)
```

## Manuscript figures

```{r}
full_median_size = 1.5

subsample_labeller <- function(subsample_level) {
  return(paste("Subsample level:", subsample_level))
}
```

### Contradictory objectives

Build plot panels (1 cohort, 1 down-sample)

```{r}

build_con_obj_plot <- function(eval_mode) {

  full_median <- median(
    filter(
      con_obj_summary_data,
      eval_mode_row == eval_mode & EVAL_MODE == "full"
    )$pop_optimal_trait_coverage
  )

  p <- con_obj_summary_data %>%
    filter(eval_mode_row == eval_mode & EVAL_MODE != "full") %>%
    ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = pop_optimal_trait_coverage,
        fill = EVAL_FIT_EST_MODE
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust=1.5
    ) +
    geom_point(
      mapping=aes(color=EVAL_FIT_EST_MODE),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      limits = c(-0.5, 50)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
    ~ evals_per_gen,
    nrow = 1,
    labeller = as_labeller(
      subsample_labeller
    )
    ) +
    labs(
      x = "Estimation mode",
      y = "Satisfactory trait coverage"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color="gray", size=2)
    )

  return(p)
}

con_obj_ds_plot <- build_con_obj_plot("down-sample")
con_obj_cohort_plot <- build_con_obj_plot("cohort")
```

Combine panels into single plot.

```{r}
# Joint title: https://wilkelab.org/cowplot/articles/plot_grid.html
con_obj_title <- ggdraw() +
  draw_label(
    "Contradictory objectives diagnostic",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

con_obj_grid <- plot_grid(
  con_obj_title,
  con_obj_ds_plot +
    labs(
      title = "Down-sampled lexicase"
    ) +
    theme(axis.title.x = element_blank()),
  con_obj_cohort_plot +
    labs(
      title = "Cohort lexicase"
    ),
  nrow = 3,
  ncol = 1,
  # align = "h",
  labels = c("", "a", "b"),
  rel_heights = c(0.075, 1, 1)
)
con_obj_grid
```

```{r}
save_plot(
  filename = paste0(plot_directory, "2023-05-10-diagnostics-con-obj-final-fig.pdf"),
  plot = con_obj_grid,
  base_width = 10,
  base_height = 8,
  dpi = 600
)
```

## Multi-path exploration

```{r}
build_explore_plot <- function(eval_mode) {

  full_median <- median(
    filter(
      explore_summary_data,
      eval_mode_row == eval_mode & EVAL_MODE == "full"
    )$max_agg_score
  )

  p <- explore_summary_data %>%
    filter(eval_mode_row == eval_mode & EVAL_MODE != "full") %>%
    ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = max_agg_score,
        fill = EVAL_FIT_EST_MODE
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust=1.5
    ) +
    geom_point(
      mapping=aes(color=EVAL_FIT_EST_MODE),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      limits = c(-0.5, 10005)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
    ~ evals_per_gen,
    nrow = 1,
    labeller = as_labeller(
      subsample_labeller
    )
    ) +
    labs(
      x = "Estimation mode",
      y = "Max aggregate score"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color="gray", size=2)
    )

  return(p)
}

explore_ds_plot <- build_explore_plot("down-sample")
explore_cohort_plot <- build_explore_plot("cohort")

explore_ds_plot
explore_cohort_plot
```

Combine panels into single plot.

```{r}
# Joint title: https://wilkelab.org/cowplot/articles/plot_grid.html
explore_title <- ggdraw() +
  draw_label(
    "Multi-path exploration diagnostic",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

explore_grid <- plot_grid(
  explore_title,
  explore_ds_plot +
    labs(
      title = "Down-sampled lexicase"
    ) +
    theme(axis.title.x = element_blank()),
  explore_cohort_plot +
    labs(
      title = "Cohort lexicase"
    ),
  nrow = 3,
  ncol = 1,
  # align = "h",
  labels = c("", "a", "b"),
  rel_heights = c(0.075, 1, 1)
)
explore_grid
```

```{r}
save_plot(
  filename = paste0(plot_directory, "2023-05-10-diagnostics-explore-final-fig.pdf"),
  plot = explore_grid,
  base_width = 10,
  base_height = 8,
  dpi = 600
)
```
