# 2023-03-18 - Exploratory experiments

Lexicase selection with phylogeny-informed evaluation on 3 diagonostics problems:

- Exploitation
- Exploration
- Contradictory objectives

```{r}
experiment_slug <- "2023-03-18-diagnostics-phylo"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)
```

## Dependencies

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(khroma)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

eval_reduction_fun <- function(eval_mode, cohort_param, ds_param) {
  if (eval_mode == "cohort") {
    return(1.0 / cohort_param)
  } else if (eval_mode == "down-sample") {
    return(ds_param)
  } else if (eval_mode == "full") {
    return(1.0)
  }
}

fit_est_mode_fun <- function(est_mode) {
  if (est_mode == "ancestor-opt") {
    return("ancestor")
  } else if (est_mode == "relative-opt") {
    return("relative")
  }
  return(est_mode)
}

summary_data$evals_per_gen <- mapply(
  eval_reduction_fun,
  summary_data$EVAL_MODE,
  summary_data$NUM_COHORTS,
  summary_data$TEST_DOWNSAMPLE_RATE
)
summary_data$evals_per_gen <- as.factor(
  summary_data$evals_per_gen
)

summary_data$DIAGNOSTIC <- as.factor(
  summary_data$DIAGNOSTIC
)
summary_data$SELECTION <- as.factor(
  summary_data$SELECTION
)
summary_data$EVAL_MODE <- as.factor(
  summary_data$EVAL_MODE
)

summary_data$EVAL_FIT_EST_MODE <- mapply(
  fit_est_mode_fun,
  summary_data$EVAL_FIT_EST_MODE
)

summary_data$EVAL_FIT_EST_MODE <- factor(
  summary_data$EVAL_FIT_EST_MODE,
  levels = c(
    "none",
    "ancestor",
    "relative"
  ),
  labels = c(
    "None",
    "Ancestor",
    "Relative"
  )
)
summary_data$NUM_COHORTS <- as.factor(
  summary_data$NUM_COHORTS
)
summary_data$TEST_DOWNSAMPLE_RATE <- as.factor(
  summary_data$TEST_DOWNSAMPLE_RATE
)
```


```{r}
# Split on diagnostic
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)
exploit_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "exploitation-rate"
)
explore_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "multipath-exploration"
)
```

Load time series data

```{r}
ts_data_loc <- paste0(working_directory, "data/time_series.csv")
ts_data <- read_csv(ts_data_loc)

ts_data$evals_per_gen <- mapply(
  eval_reduction_fun,
  ts_data$EVAL_MODE,
  ts_data$NUM_COHORTS,
  ts_data$TEST_DOWNSAMPLE_RATE
)

ts_data$EVAL_FIT_EST_MODE <- mapply(
  fit_est_mode_fun,
  ts_data$EVAL_FIT_EST_MODE
)

ts_data <- ts_data %>% mutate(
  evals_per_gen = as.factor(evals_per_gen),
  DIAGNOSTIC = as.factor(DIAGNOSTIC),
  SELECTION = as.factor(SELECTION),
  EVAL_MODE = as.factor(EVAL_MODE),
  NUM_COHORTS = as.factor(NUM_COHORTS),
  TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
  EVAL_FIT_EST_MODE = factor(
    EVAL_FIT_EST_MODE,
    levels = c(
      "none",
      "ancestor",
      "relative"
    ),
    labels = c(
      "none",
      "ancestor",
      "relative"
    )
  )
)

con_obj_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives")
exploit_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "exploitation-rate")
explore_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "multipath-exploration")
```

- max_agg_score
- num_unique_selected
- phylo_genotype_current_phylogenetic_diversity
- phylo_mean_genotype_pairwise_distance
- pop_optimal_trait_coverage

## Contradictory objectives

```{r}
contradictory_obj_final_plt <- ggplot(
    con_obj_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = pop_optimal_trait_coverage,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-final.pdf"),
  plot = contradictory_obj_final_plt,
  width = 15,
  height = 10
)
```

```{r}
contradictory_obj_pop_cov_ts <- ggplot(
    con_obj_ts_data,
    aes(
      x=ts_step,
      y=pop_optimal_trait_coverage,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) + facet_wrap(
    evals_per_gen~EVAL_MODE,
    ncol=1,
    labeller=label_both
  )
contradictory_obj_pop_cov_ts
ggsave(
  filename = paste0(plot_directory, "contra-obj-ts.pdf"),
  plot = contradictory_obj_pop_cov_ts,
  width = 10,
  height = 15
)
```

```{r}
contradictory_obj_final_mean_pw_dist_plt <- ggplot(
    con_obj_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = phylo_mean_genotype_pairwise_distance,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-phylo-mean-pw-dist.pdf"),
  plot = contradictory_obj_final_mean_pw_dist_plt,
  width = 15,
  height = 10
)
```

```{r}
contradictory_obj_phylo_mean_pw_ts <- ggplot(
    con_obj_ts_data,
    aes(
      x=ts_step,
      y=mean_genotype_pairwise_distance,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) + facet_wrap(
    evals_per_gen~EVAL_MODE,
    ncol=1,
    labeller=label_both
  ) + theme(
    legend.position="bottom"
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-ts-phylo_mean_genotype_pairwise_distance.pdf"),
  plot = contradictory_obj_phylo_mean_pw_ts,
  width = 10,
  height = 15
)
```

```{r}
contradictory_obj_ts_mrca_depth <- ggplot(
    con_obj_ts_data,
    aes(
      x=ts_step,
      y=mrca_depth,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) + facet_wrap(
    evals_per_gen~EVAL_MODE,
    ncol=1,
    labeller=label_both,
    scales = "free"
  ) + theme(
    legend.position="bottom"
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-ts-mrca_depth.pdf"),
  plot = contradictory_obj_ts_mrca_depth,
  width = 10,
  height = 15
)
```

## Multipath exploration

```{r}
explore_final_score_plt <- ggplot(
    explore_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
ggsave(
  filename = paste0(plot_directory, "explore_final_score.pdf"),
  plot = explore_final_score_plt,
  width = 15,
  height = 10
)
```

```{r}
explore_ts_max_agg_score <- ggplot(
    explore_ts_data,
    aes(
      x=ts_step,
      y=max_agg_score,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    ncol=1,
    labeller=label_both
  ) +
  theme(
    legend.position = "bottom"
  )
# explore_ts_max_agg_score
ggsave(
  filename = paste0(plot_directory, "explore-ts_max_agg_score.pdf"),
  plot = explore_ts_max_agg_score,
  width = 10,
  height = 15
)
```

```{r}
explore_final_mean_pw_dist_plt <- ggplot(
    explore_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = phylo_mean_genotype_pairwise_distance,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
ggsave(
  filename = paste0(plot_directory, "explore-phylo-mean-pw-dist.pdf"),
  plot = explore_final_mean_pw_dist_plt,
  width = 15,
  height = 10
)
```

## Exploitation rate

```{r}
exploit_final_score_plt <- ggplot(
    exploit_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
ggsave(
  filename = paste0(plot_directory, "exploit_final_score.pdf"),
  plot = exploit_final_score_plt,
  width = 15,
  height = 10
)
```

```{r}
exploit_ts_max_agg_score <- ggplot(
    exploit_ts_data,
    aes(
      x=ts_step,
      y=max_agg_score,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    ncol=1,
    labeller=label_both
  ) +
  theme(
    legend.position = "bottom"
  )
# explore_ts_max_agg_score
ggsave(
  filename = paste0(plot_directory, "exploit-ts_max_agg_score.pdf"),
  plot = exploit_ts_max_agg_score,
  width = 10,
  height = 15
)
```