# 2023-05-02 - Diagnostics experiments

```{r}
experiment_slug <- "2023-05-02-diagnostics"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)
```

## Dependencies

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read_csv(summary_data_loc)

eval_reduction_fun <- function(eval_mode, cohort_param, ds_param) {
  if (eval_mode == "cohort") {
    return(1.0 / cohort_param)
  } else if (eval_mode == "down-sample") {
    return(ds_param)
  } else if (eval_mode == "full") {
    return(1.0)
  }
}

plot_row_fun <- function(eval_mode, cohort_param, ds_param) {
  if (eval_mode == "full" && ds_param == "1") {
    return("down-sample")
  } else if (eval_mode == "full" && cohort_param == "1") {
    return("cohort")
  } else {
    return(eval_mode)
  }
}

fit_est_mode_fun <- function(est_mode) {
  if (est_mode == "ancestor-opt") {
    return("ancestor")
  } else if (est_mode == "relative-opt") {
    return("relative")
  }
  return(est_mode)
}

summary_data$evals_per_gen <- mapply(
  eval_reduction_fun,
  summary_data$EVAL_MODE,
  summary_data$NUM_COHORTS,
  summary_data$TEST_DOWNSAMPLE_RATE
)

summary_data$eval_mode_row <- mapply(
  plot_row_fun,
  summary_data$EVAL_MODE,
  summary_data$NUM_COHORTS,
  summary_data$TEST_DOWNSAMPLE_RATE
)

summary_data$EVAL_FIT_EST_MODE <- mapply(
  fit_est_mode_fun,
  summary_data$EVAL_FIT_EST_MODE
)

summary_data <- summary_data %>% mutate(
  evals_per_gen = as.factor(evals_per_gen),
  eval_mode_row = as.factor(eval_mode_row),
  DIAGNOSTIC = as.factor(DIAGNOSTIC),
  SELECTION = as.factor(SELECTION),
  EVAL_MODE = as.factor(EVAL_MODE),
  NUM_COHORTS = as.factor(NUM_COHORTS),
  TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE)
)

summary_data$EVAL_FIT_EST_MODE <- factor(
  summary_data$EVAL_FIT_EST_MODE,
  levels = c(
    "none",
    "ancestor",
    "relative"
  ),
  labels = c(
    "None",
    "Ancestor",
    "Relative"
  )
)

# Split summary data on diagnostic
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)
exploit_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "exploitation-rate"
)
explore_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "multipath-exploration"
)
```

Load time series data


```{r}
ts_data_loc <- paste0(working_directory, "data/time_series.csv")
ts_data <- read_csv(ts_data_loc)

ts_data$evals_per_gen <- mapply(
  eval_reduction_fun,
  ts_data$EVAL_MODE,
  ts_data$NUM_COHORTS,
  ts_data$TEST_DOWNSAMPLE_RATE
)

ts_data$EVAL_FIT_EST_MODE <- mapply(
  fit_est_mode_fun,
  ts_data$EVAL_FIT_EST_MODE
)

ts_data <- ts_data %>% mutate(
  evals_per_gen = as.factor(evals_per_gen),
  DIAGNOSTIC = as.factor(DIAGNOSTIC),
  SELECTION = as.factor(SELECTION),
  EVAL_MODE = as.factor(EVAL_MODE),
  NUM_COHORTS = as.factor(NUM_COHORTS),
  TEST_DOWNSAMPLE_RATE = as.factor(TEST_DOWNSAMPLE_RATE),
  EVAL_FIT_EST_MODE = factor(
    EVAL_FIT_EST_MODE,
    levels = c(
      "none",
      "ancestor",
      "relative"
    ),
    labels = c(
      "none",
      "ancestor",
      "relative"
    )
  )
)

con_obj_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "contradictory-objectives")
exploit_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "exploitation-rate")
explore_ts_data <- ts_data %>%
  filter(DIAGNOSTIC == "multipath-exploration")
```

## Contradictory objectives diagnostic

```{r}
contradictory_obj_final_plt <- ggplot(
    con_obj_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = pop_optimal_trait_coverage,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    eval_mode_row~evals_per_gen,
    # nrow=2,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color="gray", size=2)
  )
ggsave(
  filename = paste0(plot_directory, "contra-obj-final.pdf"),
  plot = contradictory_obj_final_plt + labs(title="Contradictory objectives"),
  width = 15,
  height = 10
)
```

```{r}
contradictory_obj_pop_cov_ts <- ggplot(
    con_obj_ts_data,
    aes(
      x=ts_step,
      y=pop_optimal_trait_coverage,
      fill=EVAL_FIT_EST_MODE,
      color=EVAL_FIT_EST_MODE
    )
  ) +
  stat_summary(geom="line", fun=mean) +
  stat_summary(
    geom="ribbon",
    fun.data="mean_cl_boot",
    fun.args=list(conf.int=0.95),
    alpha=0.2,
    linetype=0
  ) +
  facet_wrap(
    EVAL_MODE~evals_per_gen,
    ncol=1,
    labeller=label_both
  )
# contradictory_obj_pop_cov_ts
ggsave(
  filename = paste0(plot_directory, "contra-obj-ts.pdf"),
  plot = contradictory_obj_pop_cov_ts + labs(title="Contradictory objectives"),
  width = 10,
  height = 15
)
```

```{r}
con_obj_kw_test <- con_obj_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  kruskal_test(pop_optimal_trait_coverage ~ EVAL_FIT_EST_MODE)
kable(con_obj_kw_test)
```

```{r}
con_obj_stats <- con_obj_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  pairwise_wilcox_test(pop_optimal_trait_coverage ~ EVAL_FIT_EST_MODE) %>%
  adjust_pvalue(method = "holm")

kable(con_obj_stats)
```


## Multi-path exploration diagnostic

```{r}
explore_final_score_plt <- ggplot(
    explore_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    eval_mode_row~evals_per_gen,
    # nrow=2,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color="gray", size=2)
  )
ggsave(
  filename = paste0(plot_directory, "explore-final.pdf"),
  plot = explore_final_score_plt + labs(title="Multi-path exploration"),
  width = 15,
  height = 10
)
```

```{r}
explore_kw_test <- explore_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  kruskal_test(max_agg_score ~ EVAL_FIT_EST_MODE)
kable(explore_kw_test)
```

```{r}
explore_stats <- explore_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  pairwise_wilcox_test(max_agg_score ~ EVAL_FIT_EST_MODE) %>%
  adjust_pvalue(method = "holm")

kable(explore_stats)
```

## Exploitation rate diagnostic

```{r}
exploit_final_score_plt <- ggplot(
    exploit_summary_data,
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_grid(
    eval_mode_row~evals_per_gen,
    # nrow=2,
    labeller=label_both
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color="gray", size=2)
  )
ggsave(
  filename = paste0(plot_directory, "exploit-final.pdf"),
  plot = exploit_final_score_plt + labs(title="Exploitation rate"),
  width = 15,
  height = 10
)
```

```{r}
exploit_stats <- exploit_summary_data %>%
  filter(EVAL_MODE != "full") %>%
  group_by(EVAL_MODE, evals_per_gen) %>%
  pairwise_wilcox_test(max_agg_score ~ EVAL_FIT_EST_MODE) %>%
  adjust_pvalue(method = "holm")

kable(exploit_stats)
```

## Manuscript figures

```{r}
full_median_size = 1.5

subsample_labeller <- function(subsample_level) {
  return(paste("Subsample level:", subsample_level))
}
```

### Contradictory objectives

Build plot panels (1 cohort, 1 down-sample)

```{r}

build_con_obj_plot <- function(eval_mode) {

  full_median <- median(
    filter(
      con_obj_summary_data,
      eval_mode_row == eval_mode & EVAL_MODE == "full"
    )$pop_optimal_trait_coverage
  )

  p <- con_obj_summary_data %>%
    filter(eval_mode_row == eval_mode & EVAL_MODE != "full") %>%
    ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = pop_optimal_trait_coverage,
        fill = EVAL_FIT_EST_MODE
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust=1.5
    ) +
    geom_point(
      mapping=aes(color=EVAL_FIT_EST_MODE),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      limits = c(-0.5, 50)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
    ~ evals_per_gen,
    nrow = 1,
    labeller = as_labeller(
      subsample_labeller
    )
    ) +
    labs(
      x = "Estimation mode",
      y = "Satisfactory trait coverage"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color="gray", size=2)
    )

  return(p)
}

con_obj_ds_plot <- build_con_obj_plot("down-sample")
con_obj_cohort_plot <- build_con_obj_plot("cohort")
```

Combine panels into single plot.

```{r}
# Joint title: https://wilkelab.org/cowplot/articles/plot_grid.html
con_obj_title <- ggdraw() +
  draw_label(
    "Contradictory objectives diagnostic",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

con_obj_grid <- plot_grid(
  con_obj_title,
  con_obj_ds_plot +
    labs(
      title = "Down-sampled lexicase"
    ) +
    theme(axis.title.x = element_blank()),
  con_obj_cohort_plot +
    labs(
      title = "Cohort lexicase"
    ),
  nrow = 3,
  ncol = 1,
  # align = "h",
  labels = c("", "a", "b"),
  rel_heights = c(0.075, 1, 1)
)
con_obj_grid
```

```{r}
save_plot(
  filename = paste0(plot_directory, "2023-05-02-diagnostics-con-obj-final-fig.pdf"),
  plot = con_obj_grid,
  base_width = 10,
  base_height = 8,
  dpi = 600
)
```

## Multi-path exploration

```{r}
build_explore_plot <- function(eval_mode) {

  full_median <- median(
    filter(
      explore_summary_data,
      eval_mode_row == eval_mode & EVAL_MODE == "full"
    )$max_agg_score
  )

  p <- explore_summary_data %>%
    filter(eval_mode_row == eval_mode & EVAL_MODE != "full") %>%
    ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = max_agg_score,
        fill = EVAL_FIT_EST_MODE
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust=1.5
    ) +
    geom_point(
      mapping=aes(color=EVAL_FIT_EST_MODE),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      limits = c(-0.5, 10005)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
    ~ evals_per_gen,
    nrow = 1,
    labeller = as_labeller(
      subsample_labeller
    )
    ) +
    labs(
      x = "Estimation mode",
      y = "Max aggregate score"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color="gray", size=2)
    )

  return(p)
}

explore_ds_plot <- build_explore_plot("down-sample")
explore_cohort_plot <- build_explore_plot("cohort")

explore_ds_plot
explore_cohort_plot
```

Combine panels into single plot.

```{r}
# Joint title: https://wilkelab.org/cowplot/articles/plot_grid.html
explore_title <- ggdraw() +
  draw_label(
    "Multi-path exploration diagnostic",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

explore_grid <- plot_grid(
  explore_title,
  explore_ds_plot +
    labs(
      title = "Down-sampled lexicase"
    ) +
    theme(axis.title.x = element_blank()),
  explore_cohort_plot +
    labs(
      title = "Cohort lexicase"
    ),
  nrow = 3,
  ncol = 1,
  # align = "h",
  labels = c("", "a", "b"),
  rel_heights = c(0.075, 1, 1)
)
explore_grid
```

```{r}
save_plot(
  filename = paste0(plot_directory, "2023-05-02-diagnostics-explore-final-fig.pdf"),
  plot = explore_grid,
  base_width = 10,
  base_height = 8,
  dpi = 600
)
```

## Exploitation rate

```{r}
build_exploit_plot <- function(eval_mode) {

  full_median <- median(
    filter(
      exploit_summary_data,
      eval_mode_row == eval_mode & EVAL_MODE == "full"
    )$max_agg_score
  )

  p <- exploit_summary_data %>%
    filter(eval_mode_row == eval_mode & EVAL_MODE != "full") %>%
    ggplot(
      aes(
        x = EVAL_FIT_EST_MODE,
        y = max_agg_score,
        fill = EVAL_FIT_EST_MODE
      )
    ) +
    geom_hline(
      yintercept = full_median,
      size = full_median_size,
      alpha = 0.7,
      color = "black"
    ) +
    geom_flat_violin(
      position = position_nudge(x = .2, y = 0),
      alpha = .8,
      adjust=1.5
    ) +
    geom_point(
      mapping=aes(color=EVAL_FIT_EST_MODE),
      position = position_jitter(width = .15),
      size = .5,
      alpha = 0.8
    ) +
    geom_boxplot(
      width = .1,
      outlier.shape = NA,
      alpha = 0.5
    ) +
    scale_y_continuous(
      limits = c(5000, 10005)
    ) +
    scale_fill_bright() +
    scale_color_bright() +
    facet_wrap(
    ~ evals_per_gen,
    nrow = 1,
    labeller = as_labeller(
      subsample_labeller
    )
    ) +
    labs(
      x = "Estimation mode",
      y = "Max aggregate score"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(
        angle = 30,
        hjust = 1
      ),
      panel.border = element_rect(color="gray", size=2)
    )

  return(p)
}

exploit_ds_plot <- build_exploit_plot("down-sample")
exploit_cohort_plot <- build_exploit_plot("cohort")

exploit_ds_plot
exploit_cohort_plot
```

Combine panels into single plot.

```{r}
# Joint title: https://wilkelab.org/cowplot/articles/plot_grid.html
exploit_title <- ggdraw() +
  draw_label(
    "Exploitation rate diagnostic",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )

exploit_grid <- plot_grid(
  exploit_title,
  exploit_ds_plot +
    labs(
      title = "Down-sampled lexicase"
    ) +
    theme(axis.title.x = element_blank()),
  exploit_cohort_plot +
    labs(
      title = "Cohort lexicase"
    ),
  nrow = 3,
  ncol = 1,
  # align = "h",
  labels = c("", "a", "b"),
  rel_heights = c(0.075, 1, 1)
)
exploit_grid
```

```{r}
save_plot(
  filename = paste0(plot_directory, "2023-05-02-diagnostics-exploit-final-fig.pdf"),
  plot = exploit_grid,
  base_width = 10,
  base_height = 8,
  dpi = 600
)
```