# 2022-12-14 - Exploratory experiments

Lexicase selection with phylogeny-informed evaluation on 3 diagonostics problems:

- Exploitation
- Exploration
- Contradictory objectives

```{r}
experiment_slug <- "2022-12-14-diagnostics-lex"

working_directory <- paste0(
  "experiments/",
  experiment_slug,
  "/analysis/"
)
```

## Dependencies

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(khroma)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
print(version)
```

## Setup

```{r}
# Configure our default graphing theme
theme_set(theme_cowplot())
# Create a directory to store plots
plot_directory <- paste0(working_directory, "plots/")
dir.create(plot_directory, showWarnings=FALSE)
```

```{r}
summary_data_loc <- paste0(working_directory, "data/aggregate.csv")
summary_data <- read.csv(summary_data_loc, na.strings = "NONE")

eval_reduction_fun <- function(eval_mode, cohort_param, ds_param) {
  if (eval_mode == "cohort") {
    return(1.0 / cohort_param)
  } else if (eval_mode == "down-sample") {
    return(ds_param)
  } else if (eval_mode == "full") {
    return(1.0)
  }
}
summary_data$evals_per_gen <- mapply(
  eval_reduction_fun,
  summary_data$EVAL_MODE,
  summary_data$NUM_COHORTS,
  summary_data$TEST_DOWNSAMPLE_RATE
)
summary_data$evals_per_gen <- as.factor(
  summary_data$evals_per_gen
)

summary_data$DIAGNOSTIC <- as.factor(
  summary_data$DIAGNOSTIC
)
summary_data$SELECTION <- as.factor(
  summary_data$SELECTION
)
summary_data$EVAL_MODE <- as.factor(
  summary_data$EVAL_MODE,
)
summary_data$EVAL_FIT_EST_MODE <- factor(
  summary_data$EVAL_FIT_EST_MODE,
  levels = c(
    "none",
    "ancestor",
    "relative"
  ),
  labels = c(
    "none",
    "ancestor",
    "relative"
  )
)
summary_data$NUM_COHORTS <- as.factor(
  summary_data$NUM_COHORTS
)
summary_data$TEST_DOWNSAMPLE_RATE <- as.factor(
  summary_data$TEST_DOWNSAMPLE_RATE
)
```


```{r}
# Split on diagnostic
con_obj_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "contradictory-objectives"
)
exploit_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "exploitation-rate"
)
explore_summary_data <- filter(
  summary_data,
  DIAGNOSTIC == "multipath-exploration"
)
```

- max_agg_score
- num_unique_selected
- phylo_genotype_current_phylogenetic_diversity
- phylo_mean_genotype_pairwise_distance
- pop_optimal_trait_coverage

## Contradictory objectives

```{r}
ggplot(
    filter(
      con_obj_summary_data,
      evals_per_gen != 0.1
    ),
    aes(
      x = EVAL_FIT_EST_MODE,
      y = pop_optimal_trait_coverage,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
```

```{r}
ggplot(
    filter(
      con_obj_summary_data,
      evals_per_gen != 0.1
    ),
    aes(
      x = EVAL_FIT_EST_MODE,
      y = phylo_mean_genotype_pairwise_distance,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
```

## Multipath exploration

```{r}
ggplot(
    filter(
      explore_summary_data,
      evals_per_gen != 0.1
    ),
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
```

## Exploitation rate

```{r}
ggplot(
    filter(
      exploit_summary_data,
      evals_per_gen != 0.1
    ),
    aes(
      x = EVAL_FIT_EST_MODE,
      y = max_agg_score,
      fill = EVAL_FIT_EST_MODE
    )
  ) +
  geom_flat_violin(
    position = position_nudge(x = .2, y = 0),
    alpha = .8,
    adjust=1.5
  ) +
  geom_point(
    mapping=aes(color=EVAL_FIT_EST_MODE),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_y_continuous(
    # limits = c(-0.5, 100)
  ) +
  scale_fill_bright() +
  scale_color_bright() +
  facet_wrap(
    evals_per_gen~EVAL_MODE,
    nrow=1,
    labeller=label_both
  ) +
  theme(
    legend.position="none"
  )
```